---
title: "Bayesian GLMM Part 5"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(brms)
theme_set(theme_classic())
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
owls <- read_csv('../data/owls.csv', trim_ws=TRUE) %>%
  janitor::clean_names() %>%
  rename(food = food_treatment, sex = sex_parent)
glimpse(owls)
```


# Exploratory data analysis

Model formula:
$$
y_i \sim{} \mathcal{Pois}(\lambda_i)\\
ln(\lambda_i) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of food treatment, sex of parent, arrival time (and various interactions) on the number of sibling negotiations.
Brood size was also incorporated as an offset.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual nests.

# Fit the model


```{r fitModel, results='markdown', eval=FALSE, hidden=TRUE}
## Amount of Sibling negotiation (vocalizations when parents are absent)
## Foot treatment (deprived or satiated
## Sex of parent
## Arrival time of parent
## nest as random
## Brood size offset
owls = owls %>% mutate(nest = factor(nest),
                       food = factor(food),
                       sex = factor(sex),   
                       ncalls = SiblingNegotiation)
ggplot(data = owls) +
    geom_boxplot(aes(y = ncalls, x = nest)) + 
    facet_grid(sex ~ food)

ggplot(data=owls) +
  geom_boxplot(aes(y=ncalls/brood_size,  x=sex,  color=food)) +
  facet_wrap(~nest)

ggplot(data = owls,aes(y = ncalls/brood_size, x = arrival_time, color=sex)) +
    geom_point() + 
    geom_smooth(method='lm') +
  facet_grid(~food)

ggplot(data = owls,aes(y = ncalls, x = arrival_time, color=food)) +
    geom_point() + 
    geom_smooth(method='lm') 
#  facet_grid(~food)

ggplot(data = owls,aes(y = NegPerChick, x = arrival_time, color=food)) +
    geom_point() + 
  geom_smooth(method='lm') +
  facet_wrap(~nest,  scale='free_y')



owls_rstanP <- stan_glmer(ncalls ~ food+sex +
                           offset(log(brood_size)) + (1|nest),
                         dat=owls,  family=poisson(link='log'), refresh=0, 
                         iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)


owls_rstanP <- stan_glmer(ncalls ~ food+scale(arrival_time) +
                           offset(log(brood_size)) + (1|nest),
                         dat=owls,  family=poisson(link='log'), refresh=0, 
                         iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_rstanP %>% get_variables()
plot(owls_rstanP,  'mcmc_trace', regex_pars='^.Intercept|Food|Arrival|[sS]igma')
plot(owls_rstanP,  'mcmc_acf_bar', regex_pars='^.Intercept|Food|Arrival|[sS]igma')
plot(owls_rstanP,  'mcmc_rhat_hist', regex_pars='^.Intercept|Food|Arrival|[sS]igma')
plot(owls_rstanP,  'mcmc_neff_hist', regex_pars='^.Intercept|Food|Arrival|[sS]igma')


preds <- posterior_predict(owls_rstanP,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median))
plot(owls_resids)


owls_rstanNB <- stan_glmer(ncalls ~ food+sex +
                           offset(log(brood_size)) + (1|nest),
                         dat=owls,  family=neg_binomial_2(link='log'), refresh=0, 
                         iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_rstanNB <- stan_glmer(ncalls ~ food+scale(arrival_time) +
                           offset(log(brood_size)) + (1|nest),
                         dat=owls,  family=neg_binomial_2(link='log'), refresh=0, 
                         iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_rstanNB %>% get_variables()
plot(owls_rstanNB,  'mcmc_trace', regex_pars='^.Intercept|Food|Arrival|[sS]igma')
plot(owls_rstanNB,  'mcmc_acf_bar', regex_pars='^.Intercept|Food|Arrival|[sS]igma')
plot(owls_rstanNB,  'mcmc_rhat_hist', regex_pars='^.Intercept|Food|Arrival|[sS]igma')
plot(owls_rstanNB,  'mcmc_neff_hist', regex_pars='^.Intercept|Food|Arrival|[sS]igma')


preds <- posterior_predict(owls_rstanNB,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median))
plot(owls_resids)
testZeroInflation(owls_resids)

testTemporalAutocorrelation(owls_resids,  time=owls$arrival_time)
owls_resids1 <- recalculateResiduals(owls_resids,  group=interaction(owls$arrival_time,  owls$nest),  aggregateBy = mean)
testTemporalAutocorrelation(owls_resids1,  time=unique(owls$arrival_time))

## Cant use zero inflation with glmer

owls_rstan1 <- stan_glmer(ncalls ~ food+scale(arrival_time) +
                           offset(log(brood_size)) + (food*scale(arrival_time)|nest),
                         dat=owls,  family=neg_binomial_2(link='log'), refresh=0, 
                         iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_rstan1 %>% get_variables()
plot(owls_rstan1,  'mcmc_trace', regex_pars='^.Intercept|^Food|^scale.Arrival|[sS]igma')
plot(owls_rstan1,  'mcmc_acf_bar', regex_pars='^.Intercept|^Food|^scale.Arrival|[sS]igma')
plot(owls_rstan1,  'mcmc_rhat_hist', regex_pars='^.Intercept|^Food|^scale.Arrival|[sS]igma')
plot(owls_rstan1,  'mcmc_neff_hist', regex_pars='^.Intercept|^Food|^scale.Arrival|[sS]igma')

preds <- posterior_predict(owls_rstan1,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median))
plot(owls_resids)

```

```{r fitModel.brms, results='markdown', eval=FALSE, hidden=TRUE}
## Amount of Sibling negotiation (vocalizations when parents are absent)
## Foot treatment (deprived or satiated
## Sex of parent
## Arrival time of parent
## nest as random
## Brood size offset
owls = owls %>% mutate(nest =factor(nest),
                       food = factor(food),
                       sex = factor(sex),   
                       ncalls = SiblingNegotiation)
ggplot(data = owls) +
    geom_boxplot(aes(y = ncalls, x = nest)) + 
    facet_grid(sex ~ food)

ggplot(data=owls) +
  geom_boxplot(aes(y=ncalls,  x=sex,  color=food)) +
  facet_wrap(~nest)

ggplot(data = owls,aes(y = ncalls, x = arrival_time, color=sex)) +
    geom_point() + 
    geom_smooth(method='lm') +
  facet_grid(~food)

owls_form <- bf(ncalls ~ food+scale(arrival_time) +
                  offset(log(brood_size)) + (1|nest),
                family=poisson(link='log'))

 owls_form <- bf(ncalls ~ food+sex +
                  offset(log(brood_size)) + (1|nest),
                family=(link='log'))
owls_brmsP <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_brmsP %>% get_variables()
mcmc_plot(owls_brmsP,  type='trace')
mcmc_plot(owls_brmsP,  type='acf_bar')
mcmc_plot(owls_brmsP,  type='rhat_hist')
mcmc_plot(owls_brmsP,  type='neff_hist')
mcmc_plot(owls_brmsP,  type='trace', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsP,  type='acf_bar', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsP,  type='rhat_hist', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsP,  type='neff_hist', regex_pars='^b.Intercept|Food|Arrival|sd')

library(DHARMa)
preds <- posterior_predict(owls_brmsP,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median),
                            integerResponse = TRUE)
plot(owls_resids)
testZeroInflation(owls_resids)



owls_form <- bf(ncalls ~ food+sex +
                  offset(log(brood_size)) + (1|nest),
                family=negbinomial(link='log'))
owls_brmsNB <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

mcmc_plot(owls_brmsNB,  type='trace')
mcmc_plot(owls_brmsNB,  type='acf_bar')
mcmc_plot(owls_brmsNB,  type='rhat_hist')
mcmc_plot(owls_brmsNB,  type='neff_hist')
preds <- posterior_predict(owls_brmsNB,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median),
                            integerResponse = TRUE)
plot(owls_resids)
testZeroInflation(owls_resids)

owls_form <- bf(ncalls ~ food+scale(arrival_time) +
                  offset(log(brood_size)) + (1|nest),
                zi ~ 1, 
                family=zero_inflated_poisson(link='log'))

owls_brmsZIP <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_brmsZIP %>% get_variables()

mcmc_plot(owls_brmsZIP,  type='trace')
, regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZIP,  type='acf_bar')
, regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZIP,  type='rhat_hist')
, regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZIP,  type='neff_hist')
, regex_pars='^b.Intercept|Food|Arrival|sd')


preds <- posterior_predict(owls_brmsZIP,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median))
plot(owls_resids)


owls_form <- bf(ncalls ~ food+scale(arrival_time) +
                  offset(log(brood_size)) + (1|nest),
                zi ~ scale(arrival_time), 
                family=zero_inflated_poisson(link='log'))

owls_brmsZIP1 <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_brmsZIP1 %>% get_variables()
mcmc_plot(owls_brmsZIP1,  type='trace', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZIP1,  type='acf_bar', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZIP1,  type='rhat_hist', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZIP1,  type='neff_hist', regex_pars='^b.Intercept|Food|Arrival|sd')


preds <- posterior_predict(owls_brmsZIP1,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median),
                            integerResponse=TRUE)
plot(owls_resids)


owls_form <- bf(ncalls ~ food+scale(arrival_time) +
                  offset(log(brood_size)) + (1|nest),
                zi ~ scale(arrival_time), 
                family=zero_inflated_negbinomial(link='log'))

owls_brmsZINB <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_brmsZINB %>% get_variables()
mcmc_plot(owls_brmsZINB,  type='trace', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZINB,  type='acf_bar', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZINB,  type='rhat_hist', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZINB,  type='neff_hist', regex_pars='^b.Intercept|Food|Arrival|sd')


preds <- posterior_predict(owls_brmsZINB,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median))
plot(owls_resids)


owls_form <- bf(ncalls ~ food*scale(arrival_time) +
                  offset(log(brood_size)) + (scale(arrival_time)|nest),
                zi ~ scale(arrival_time), 
                family=zero_inflated_negbinomial(link='log'))

owls_brmsZINB2 <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)

owls_brmsZINB1 %>% get_variables()
mcmc_plot(owls_brmsZINB2,  type='trace', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZINB1,  type='acf_bar', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZINB1,  type='rhat_hist', regex_pars='^b.Intercept|Food|Arrival|sd')
mcmc_plot(owls_brmsZINB1,  type='neff_hist', regex_pars='^b.Intercept|Food|Arrival|sd')


preds <- posterior_predict(owls_brmsZINB2,  nsamples=250,  summary=FALSE)
owls_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = owls$ncalls,
                            fittedPredictedResponse = apply(preds, 2, median),
                            integerResponse = TRUE)
plot(owls_resids)

testZeroInflation(owls_resids)
testDispersion(owls_resids)

owls_form <- bf(ncalls ~ food*sex +
                  offset(log(brood_size)) + (1|nest),
                zi ~ food+sex, 
                family=zero_inflated_negbinomial(link='log'))

owls_brmsZINB2 <- brm(owls_form,  data=owls, refresh=0, 
                  iter=5000,  warmup=2000,  thin=10,  chains=3, cores=3)
testTemporalAutocorrelation(owls_resids,  time=owls$arrival_time)
owls_resids1 <- recalculateResiduals(owls_resids,  group=interaction(owls$arrival_time,  owls$nest),  aggregateBy = mean)
testTemporalAutocorrelation(owls_resids1,  time=unique(owls$arrival_time))

summary(owls_brmsZINB2)
tidyMCMC(owls_brmsZINB2$fit,  estimate.method='median',
         conf.int=TRUE,  conf.method='HPDinterval',
         rhat=TRUE,  ess=TRUE)


owls_grid = with(owls, list(food=levels(food),
                            arrival_time=modelr::seq_range(arrival_time,  n=100)
                            ))
owls_grid = with(owls,  list(food=levels(food),
                             arrival_time=seq(min(arrival_time),  max(arrival_time),  len=100)))

newdata = emmeans(owls_brmsZINB2, ~sex|food, at=owls_grid,
                  type='response') %>%
  as.data.frame
head(newdata)
ggplot(newdata) +
  geom_pointrange(aes(y=prob,  x=food,  color=sex,  ymin=lower.HPD,  ymax=upper.HPD),  position=position_dodge(width=0.2))
ggplot(newdata, aes(y=prob, x=as.numeric(sex)) +
    geom_line(aes(color=food)) +
    geom_ribbon(aes(ymin=lower.HPD, ymax=upper.HPD, fill=food), alpha=0.2)

```
