---
title: "Bayesian GLMM Part 4"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(tidybayes)
require(brms)
theme_set(theme_classic())
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
mckeon <- read_csv('../data/mckeon.csv', trim_ws=TRUE) %>%
  janitor::clean_names() %>% 
  mutate(block = factor(block),
         symbiont = factor(symbiont, levels=c('none','crabs','shrimp','both')))
glimpse(mckeon)
```


# Exploratory data analysis

Model formula:
$$
y_i \sim{} \mathcal{N}(n, p_i)\\
ln\left(\frac{p_i}{1-p_1}\right) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of symbionts on the probability of the colony experiencing predation.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual coral colonies.

# Fit the model

```{r fitModel.brms, results='markdown', eval=FALSE, hidden=TRUE}
ggplot(mckeon, aes(y=predation, x=symbiont)) +
    geom_point(position=position_jitter(width=0.2, height=0))+
    facet_wrap(~block)

ggplot(mckeon, aes(y=predation, x=symbiont)) +
    geom_point()+
    facet_wrap(~block)
```

## Fit the model

Run the model without any priors to see what it gives us!
```{r, eval=F}
mckeon_form <- bf(predation | trials(1) ~ symbiont + (1|block),
                  family=binomial(link='logit'))

mckeon_brms <- brm(mckeon_form,
                   data=mckeon,
                   save_all_pars = TRUE,
                   sample_prior = "only",
                   iter=5000, warmup=2000, chains=3, thin=5, refresh=0,
                   cores=3)
prior_summary(mckeon_brms)

```

Interestingly, it doesn't pick very good priors for this distribution, so let's fit it with new priors...

```{r}
priors <- 
  prior(normal(0,2.5), class = "Intercept") + # set intercept in middle (plogis(0) = 0.5)
  prior(normal(0,1), class = "b") +
  prior(gamma(2,1), class = "sd")
standist::visualize("normal(0, 2.5)", "normal(0,1), gamma(2,1)") # on the link scale


mckeon_form1 <- bf(predation | trials(1) ~ symbiont + (1|block),
                  family=binomial(link='logit'))
mckeon_brm1 <- brm(mckeon_form1,
                   data = mckeon,
                   prior = priors,
                   save_all_pars = TRUE,
                   sample_prior = "yes",
                   iter=5000, warmup=2500, chains=3, thin=5,
                   cores=3)

mckeon_form2 <- bf(predation | trials(1) ~ symbiont + (symbiont|block),
                  family=binomial(link='logit'))
mckeon_brm2 <- brm(mckeon_form2,
                   data = mckeon,
                   prior = priors,
                   save_all_pars = TRUE,
                   sample_prior = "yes",
                   iter=5000, warmup=2500, chains=3, thin=5,
                   cores=3)

l1 <- loo(mckeon_brm1); l2 <- loo(mckeon_brm2)
loo_compare(l1, l2)
```
Second model is better!
`r -7.5*-2` DIC, much greater than the SE!

```{r}
(pars <- mckeon_brm2 %>% get_variables())
wch <- grepl("^b_.*|[Ss]igma|^sd|^cor_.*", pars, perl=T)
pars[wch] # return specific columns only

hypothesis(mckeon_brm2, "symbiontcrabs = 0") %>% plot()
```

Problem: the posterior seems entirely driven by the prior, which is not good...
Prior for slope might be too small, try 5 instead of 1 for variance.
```{r}
priors <- 
  prior(normal(0,2.5), class = "Intercept") + # set intercept in middle (plogis(0) = 0.5)
  prior(normal(0,5), class = "b") +
  prior(gamma(2,1), class = "sd")

mckeon_brm2 <- mckeon_brm2 %>% update(prior = priors)
hypothesis(mckeon_brm2, "symbiontcrabs = 0") %>% plot()
```

Looks much better!

Notice that you cannot examine your prior distribution for the intercept!
```{r}
hypothesis(mckeon_brm2, "Intercept = 0") %>% plot()
hypothesis(mckeon_brm2, "prior_Intercept = 0", class="") %>% plot()
```
Can plot these two, but problem is that they are not a fair comparison! We don't scale our variables, however, the intercept shown here is the intercept after scaling... So brms doesn't want to give us back the intercept that was used in the model (since the intercept was scaled in the first place!), so in a way is saving us from making a mistake. However, since we are using a categorical variable, there was no scaling done, so we actually CAN do this comparison in this instance!

## Mixing diagnostics

```{r}
mcmc_plot(mckeon_brm2,  type='trace', regex_pars='^.Intercept|symbiont|sd')
mcmc_plot(mckeon_brm2,  type="acf_bar", regex_pars='^.Intercept|symbiont|sd')
mcmc_plot(mckeon_brm2,  type='rhat_hist', regex_pars='^.Intercept|symbiont|sd')
mcmc_plot(mckeon_brm2,  type='neff_hist', regex_pars='^.Intercept|symbiont|sd')

# DHARMa:
preds <- posterior_predict(mckeon_brm2,  nsamples=250,  summary=FALSE)
mckeon_resids <- createDHARMa(simulatedResponse = t(preds),
                            observedResponse = mckeon$predation,
                            fittedPredictedResponse = apply(preds, 2, median),
                            integerResponse=TRUE)
plot(mckeon_resids)
```
Note that we are slightly overdispersed, but it is actually difficult to conceive overdispersed binomial models, but we could use beta-binomial if we had lots of data or if it were a true binomial model with multiple events rather than a bernoulli/logistic model.

# Partial plots

```{r}
mckeon_brm2 %>%
  conditional_effects() %>%
  plot(points = T)
ggpredict(mckeon_brm2) %>% plot
```

# Model interpretation
```{r}
summary(mckeon_brm2)
# tidyMCMC(mckeon_brm2, conf.method = "HPDinterval", conf.int = T)
```

```{r}
mckeon_brm2 %>% bayes_R2(re.form=NA, summary=FALSE) %>% median_hdci()
```
`re.form` forces it to marginalize based on the 'random' effects. By setting it to NA, we are not conditioning on any random effects.

```{r}
mckeon_brm2 %>% bayes_R2(re.form=~(1|block), summary=FALSE) %>% median_hdci()
```
This R^2 is now conditional on the random effect of block, and thus includes more sources of variance.

```{r}
mckeon_brm2 %>% bayes_R2(re.form=~(symbiont|block), summary=FALSE) %>% median_hdci()
```
This R^2 takes into account the individual slope responses to each symbiont.


# Predictions

```{r}
mckeon_brm2 %>%
  emmeans(~symbiont, type = "response") %>%
  pairs()

# Returning the probability of surviving
mckeon_brm2 %>%
  emmeans(~symbiont, type = "link") %>%
  pairs() %>%
  gather_emmeans_draws() %>%
  mutate(fit = plogis(.value)) %>%
  summarise(P = sum(fit > 0.5)/n())

# Same, but returning odds ratios:
mckeon_brm2 %>%
  emmeans(~symbiont, type = "link") %>%
  pairs() %>%
  gather_emmeans_draws() %>%
  mutate(fit = exp(.value))

# Planned contrasts:
cmat=cbind(
    crab_vs_shrimp= c(0,   1,  -1,   0),
    one_vs_both   = c(0,-1/2,-1/2,   1),
    symbiont      = c(1,-1/3,-1/3,-1/3)
)

mckeon_brm2 %>%
  emmeans(~symbiont, type = "link") %>%
  contrast(method = list(cmat)) %>%
  gather_emmeans_draws() %>%
  mutate(fit = plogis(.value)) %>%
  median_hdci(fit)
```




# Extra code already in the doc

```{r, eval=F}

#prior_summary(mckeon_brms1)
#posterior_vs_prior(mckeon_brms1, color_by='vs', group_by=TRUE,
#                   facet_args=list(scales='free_y'))

ggpredict(mckeon_brm2) %>% plot

mckeon_brms1 %>% get_variables()
nms=colnames(as.matrix(mckeon_brms1))
wch = grep('^b.Intercept|^b.symbiont|^sd',nms)
#posterior_vs_prior(mckeon_brms1, color_by='vs', group_by=TRUE,
#                   facet_args=list(scales='free_y'), pars=nms[wch])



summary(mckeon_brms1)
tidyMCMC(mckeon_brms1$fit,  estimate.method = 'median',
         conf.int=TRUE,  conf.method = 'HPDinterval',
         rhat=TRUE, ess=TRUE)
tidyMCMC(mckeon_brms1$fit,conf.int=TRUE, conf.method='HPDinterval',
         rhat=TRUE, ess=TRUE,  pars=nms[wch])
tidyMCMC(mckeon_brms1$fit,conf.int=TRUE, conf.method='HPDinterval',
         rhat=TRUE, ess=TRUE)

emmeans(mckeon_brms1, pairwise~symbiont, type='response')
mckeon_em = emmeans(mckeon_brms2, pairwise~symbiont, type='link')$contrasts %>%
      gather_emmeans_draws() %>%
      mutate(PEff=exp(.value))#,
             #Prob = plogis(.value))

mckeon_em %>%
  group_by(contrast) %>%
  dplyr::select(contrast, PEff) %>%
  median_hdi

mckeon_em %>%
  group_by(contrast) %>%
  summarize(Prob=sum(PEff>1)/n())


mckeon_em %>%
  group_by(contrast) %>%
  median_hdi(Fit)

mckeon_em %>%
  group_by(contrast) %>%
  summarize(sum(Fit>1)/n())                                                                      

newdata = emmeans(mckeon_brms1, ~symbiont, type='response') %>% as.data.frame
head(newdata)
ggplot(newdata, aes(y=prob, x=symbiont)) +
    geom_pointrange(aes(ymin=lower.HPD, ymax=upper.HPD))

```

# References
