---
title: "Data wrangling"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse) #for data wrangling
```

Link to the data transformation cheatsheet 

https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf

Important data manipulation functions:

+--------------------------+-----------------------------------+------------+
|Task                      |Function                           |Package     |
+==========================+===================================+============+
|Sorting                   |`arrange()`                        |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Adding columns            |`mutate()`                         |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Transformations           |`mutate()`                         |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Re-ordering factor levels |`factor(,levels=)`                 |base        |
+--------------------------+-----------------------------------+------------+
|Re-labelling              |`factor(,lab=)`                    |base        |
+--------------------------+-----------------------------------+------------+
|                          |`recode()`                         |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Re-naming columns         |`rename(,replace=)`                |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Filtering/Subsetting      |indexing                           |base        |
+--------------------------+-----------------------------------+------------+
|~ columns                 |`select(,...)`                     |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|                          |`pull(,...)`                       |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|~ rows                    |`filter(,...)`                     |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Unique combinations       |`distinct()`                       |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Reshaping data            | `pivot_longer()`, `pivot_wider()` | **tidyr**  |
+--------------------------+-----------------------------------+------------+
|Split/combine columns     | `separate()`, `unite()`           | **tidyr**  |
+--------------------------+-----------------------------------+------------+
|Aggregating               |`group_by()` `summarise()`         |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|                          |`group_by()` `count()`             |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Merging/joining           |`*_join()`                         |**dplyr**   |
+--------------------------+-----------------------------------+------------+
|Extracting data structure |`expand()`                         |**tidyr**   |
+--------------------------+-----------------------------------+------------+
|                          |`crossing()`                       |**tidyr**   |
+--------------------------+-----------------------------------+------------+


# Piping

# Data files
 
```{r getData, results='markdown', eval=TRUE}
load(file='../data/manipulationDatasets.RData')
dat.1 %>% head
```

Sorting data
================
```{r, results='markup'}
# Arrange by the Treatment, then Time
dat.1 %>% 
  arrange(Treatment, Time)

# Arrange by Treatment, then by the mean of Resp 1 and Resp2
dat.1 %>% 
  mutate(Resp = mean(c(Resp1, Resp2))) %>%
  arrange(Treatment, Resp)
```

Adding columns - mutate
===========================

```{r}
dat.1 %>%
  mutate(
    Dose = fct_relevel(Dose, c('L', 'M', 'H')),
    Dose = fct_recode(Dose, High = "H", Medium = "M")) %>%
  as_tibble %>%
  pull(Dose)


dat.1 %>% 
  mutate(leadResp2 = lead(Resp2), 
         lagResp1 = lag(Resp1))

dat.1 %>%
  mutate(rank_time = min_rank(Time),
         dens_rank_time = dense_rank(Time))

dat.1 %>%
  mutate(row_Resp1 = row_number(Resp1), 
         row_time = row_number(Time),
         rank_time = min_rank(Time),
         ntile_Resp1 = ntile(Resp1, 4),
         medium_Resp1 = between(Resp1, 20, 40),
         fResp1A = case_when(Resp1 < 31 ~ "Low",
                 between(Resp1, 31, 50) ~ "Medium",
                 Resp1 > 50 ~ "High"),
         fResp1B = cut(Resp1, breaks = c(0, 31, 50, 200),
                              labels = c("Low", "Medium", "High")),
         fResp1C = cut(Resp1, breaks = 2,
                              labels = c("Low", "High")))
```


Summarising (aggregating) data
=================================
```{r}
SE <- function(x) sd(x) / sqrt(length(x))
dat.1 %>% 
  summarise(MeanResp1 = mean(Resp1), VarResp1 = var(Resp1),
            SEM = SE(Resp1))

# Using anonymous function:
dat.1 %>% summarise(MeanResp1 = mean(Resp1), VarResp1 = var(Resp1),
          SEM = (function(x) sd(x) / sqrt(length(x)))(Resp1))

# Summarize across a number of variables
dat.1 %>% summarise(across(c(Resp1, Resp2), 
                           list(Mean = mean, Var = var)))

# Get mean and var
dat.1 %>% summarise(across(where(is.numeric), 
                           list(Mean = mean, Var = var)))

# Get mean and length
dat.1 %>% summarize(across(where(is.numeric),  mean),
          across(where(is.factor),  length))

Variable <- c("Resp1", "Resp2")
dat.1 %>% summarise(across(all_of(Variable), list(Mean = mean, Var = var)))
```


Grouping (=aggregating)
=========================
```{r}
dat.1 %>%
  arrange(Resp1) %>%
    group_by(Treatment, Plot) %>%
    summarise(Mean = mean(Resp1),
              Var = var(Resp1),
              N = n(),
              First = first(Resp1))
```

```{r}
tikus[1:10, c(1:3, 76:77)]

tikus %>% 
  # head(10) %>%
  # select("Psammocora contigua", "Psammocora digitata", "Pocillopora damicornis", "time", "rep") %>%
  arrange(`Pocillopora damicornis`) %>%
  group_by(time) %>%
  summarise(across(all_of(c("Psammocora contigua", "Psammocora digitata", "Pocillopora damicornis")), list(Mean = mean))) %>%
  ungroup()

```


Subset columns
=================
```{r}
dat.1 %>% dplyr::select(-starts_with(c("T","D")))

dat.1 %>% dplyr::select(contains("e"))

dat.1 %>% dplyr::select(where(is_numeric))

dat.1 %>% select(matches("^.{4}$")) # regex for get only columns with 4-characters

```


## Regular expressions (regex)
```{r}
nasa %>% 
  select(lat, long, starts_with("cloud"))

nasa %>% 
  select(lat, long, matches("^cloud.*$")) %>% head
```

```{r}
# Select rep, time, but not any with pora
tikus %>% select(rep, time, !contains("pora")) %>% 
  ncol()

# Same thing, but older way of doing it
tikus %>% select(rep, time, everything(), -contains("pora")) %>%
  ncol()
```

```{r}
dat.1 %>% rename(Exposure = Treatment, Richness = Resp1) %>%
  relocate(Richness, .before=Plot)
```

https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf

Filtering
=============
```{r}
dat.1 %>% filter(Dose == "H" & Resp1 < 25) %>% nrow
dat.1 %>% filter(Dose == "H" | Resp1 < 25) %>% nrow
```

Be careful about not being ambiguous!
```{r}
dat.1 %>%
  filter(Resp1 < 40 & (Time > 1 | Dose == "L")) %>% nrow
dat.1 %>%
  filter((Resp1 < 40 & Time > 1) | Dose == "L") %>% nrow
```

```{r}
nasa %>% 
  dplyr::filter(month == 2, year == max(year)) %>%
  filter(ozone == max(ozone))
nasa %>% 
  dplyr::filter(month == 2 & year == max(year)) %>%
  arrange(-ozone) %>% slice(1)

```

```{r}
nasa %>% filter(month == 1, year == max(year)) %>%
  filter(ozone > 320 & ozone < 325)
nasa %>% 
  filter(month == 1, year == max(year), between(ozone, 320, 325))
```

Always make sure to drop 'ghost' levels
```{r}
dat.1 %>%
  filter(Plot == "P1") %>%
  pull(Plot) %>%
  levels
dat.1 %>%
  filter(Plot == "P1") %>%
  pull(Plot) %>%
  droplevels()
```


Reshaping data
=================

## Pivot longer

```{r}
data.w %>%
  pivot_longer(cols = starts_with("Time"),
               names_to = "time", values_to = "min") %>%
  mutate(time = as.numeric(str_extract(time, "(?<=Time.).")))

# easier way:
data.w %>%
  pivot_longer(cols = starts_with("Time"),
               names_to = "time", values_to = "min",
               names_prefix = "Time.") %>% # makes it way easier!!
  mutate(time = as.numeric(time))
```


## Pivot wider

```{r}
data %>%
  pivot_wider(names_from = Within, values_from = c(Resp1, Resp2))
```

Combining data
=================

```{r}
data.bio <- data.bio %>% relocate(Resp1, Resp2, .after=Subplot)

data.bio %>% full_join(data.chem) %>%
  arrange(Plot, Subplot)

data.bio %>% left_join(data.chem) %>%
  arrange(Plot, Subplot)

data.bio %>% right_join(data.chem) %>%
  arrange(Plot, Subplot)

data.bio %>% inner_join(data.chem) %>%
  arrange(Plot, Subplot)

data.bio %>% rbind(data.bio[6:9,]) %>%
  left_join(data.chem) %>%
  arrange(Plot, Subplot) # same as vlookup!

```


Applied examples
===================

```{r}

summary_tikus <- tikus %>%
  rename(`Acropora aspera` = `Acropera aspera`) %>%
  select(time, rep, starts_with("Acropora")) %>%
  pivot_longer(cols = c(-rep, -time), names_to = "species", values_to = "abundance") %>%
  mutate(cover = abundance / (10*100),
         per_cover = cover * 100) %>%
  group_by(time, rep) %>%
  summarise(total_acropora_cover = sum(cover)) %>%
  ungroup() %>%
  group_by(time) %>% # averaging across all sites now
  summarise(mean_yearly_acropora_cover = mean(total_acropora_cover),
            sd_yearly_acropora_cover = sd(total_acropora_cover)) %>%
  ungroup() %>%
  mutate(year = 1900 + as.numeric(as.character(time)))
  

```

```{r}
summary_tikus <- 
  tikus %>%
  rename(`Acropora aspera` = `Acropera aspera`) %>%
  # select(time, rep, starts_with("Acropora")) %>%
  pivot_longer(cols = c(-rep, -time), names_to = "species", values_to = "abundance") %>%
  mutate(cover = abundance / (10*100),
         per_cover = cover * 100) %>%
  separate(col = species, c("genus", "species")) %>%
  group_by(genus, time, rep) %>%
  summarise(total_cover = sum(cover)) %>%
  ungroup() %>%
  mutate(genus = fct_lump_n(genus, 5, total_cover)) %>%
  group_by(genus, time) %>% # averaging across all sites now
  summarise(mean_yearly_cover = mean(total_cover),
            sd_yearly_cover = sd(total_cover)) %>%
  ungroup() %>%
  mutate(year = 1900 + as.numeric(as.character(time)))
```

```{r}
theme_set(theme_classic())
summary_tikus %>%
   ggplot(aes(x = year, y = mean_yearly_cover, group = genus)) +
  facet_wrap(~genus) +
   geom_line(linetype = "dashed") +
  geom_area(aes(fill = genus), alpha = 0.5) +
   labs(x = "Year", y = "Mean Acropora Cover") +
   scale_y_continuous(labels = scales::label_percent(),
                      expand = expansion()) +
  scale_x_continuous(expand = expansion())
```



