---
title: "GLMM example 5"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations
 
Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(broom.mixed) #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(ggeffects) #for effects plots in ggplotjk
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(DHARMa)    #for assessing dispersion etc
library(glmmTMB)    #for glmmTMB
library(performance) #for diagnostic plots
library(see)         #for diagnostic plots
library(lme4)       #for glmer
library(glmmTMB)    #for glmmTMB
theme_set(theme_classic())
```

# Scenario

Some ornithologists were interested in the degree of sibling negotiations in owl
chicks.  Specifically, they wanted to explore how sibling negotiations were
affected by feeding satiety and the sex of the parent returning to the nest.
The ornithologists had access to a number of owl nests and were able to count
(via recording equipment) the number of sibling negotiations (calls) that the
owl chicks made when the parent returned to the nest.

We could hypothesise that the chicks might call more if they were hungry.  As
part of the investigation, the researchers were able to provided supplimentary
food.  As such, they were able to manipulate the conditions such that sometimes
the chicks in a nest would be considered deprived of supplimentary food and at
other times they were satiated.  

As a parent returned, the researchers recorded the number of sibling
negotiations (calls) along with the sex of the parent.  Since the number of
calls is likely to be a function of the number of chicks (the more chicks the
more calls), the researchers also counted the number of siblings in the brood. 

Each nest was measured on multiple occasions.  Hence, we must include the nest
as a random effect to account for the lack of independence between observations
on the same set of siblings.

# Read in the data

```{r readData, results='markdown', eval=TRUE}
owls <- read_csv('../data/owls.csv', trim_ws=TRUE) %>%
  janitor::clean_names()
  
glimpse(owls)
```

# Data preparation
Sibling negotiation is the total number of calls, our y-var
Negotiations per chick is similarly a y-variable, but not as easy to model.

This is similar to a density problem - brood size can be used as an offset on the total number of sibling negotiations, thus we don't have to resort to a gamma model.

Note that brood size should go above each nest, as it applies to every nest the same
The next factor down is nest proper (a random effect)
Next is food, either the nests were deprived or not deprived (fixed effect)
Finally, the sex of the parent
Followed by the arrival time of the parent at that nest on that occasion

```{r}
owls <- owls %>%
  mutate(nest = factor(nest), 
         food = factor(food_treatment),
         sex = factor(sex_parent),
         ncalls = sibling_negotiation)
```


# Exploratory data analysis

```{r}
owls %>%
  ggplot(aes(x = food, y = ncalls, col = sex)) +
  geom_violin() +
  geom_point(position=position_jitterdodge(jitter.height=0, dodge.width=0.9)) 
  # geom_point(position = ggforce::position_jitternormal(sd_y = 0, sd_x = 0.1))
```

```{r}
owls %>%
  ggplot(aes(x = ncalls)) +
  geom_histogram() +
  facet_grid(food ~ sex)
```
Tons of zeros for the number of calls in satiated individuals - clearly satiated is important!

If we want to use offset of brood size, need to show this 1:1 relationship!

```{r}
owls %>% 
  ggplot(aes(y = ncalls, x = brood_size, color = sex)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  facet_grid(~food) + 
  scale_x_log10() +
  scale_y_continuous(trans = scales::pseudo_log_trans()) +
  geom_abline(aes(intercept = 0, slope = 1), linetype = "dashed")
```

`scales::pseudo_log_trans()` is good for dealing with zero-inflated values and still log it!

**NB:** We will need to use log(brood_size to ensure it is 1:1 with ncalls (which when using a Poisson, uses a log-link)). 

```{r}
owls %>% 
  ggplot() +
  geom_point(aes(y=ncalls, x = food, color = sex), position = position_dodge(0.5)) +
  facet_wrap(~nest)
```


Model formula:
$$
y_i \sim{} \mathcal{Pois}(\lambda_i)\\
ln(\lambda_i) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of food treatment, sex of parent, arrival time (and various interactions) on the number of sibling negotiations.
Brood size was also incorporated as an offset.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual nests.



# Fit the model {.tabset .tabset-faded}
Because we know we are interested in any interactions between food and sex, let us add this in already.
```{r}
owls_glmm1 <- glmmTMB(ncalls ~ food * sex + offset(log(brood_size)) + (1|nest), family = poisson(link = "log"), 
        data = owls, REML=F)
owls_glmm2 <- glmmTMB(ncalls ~ food + sex + offset(log(brood_size)) + (1|nest), family = poisson(link = "log"), 
        data = owls, REML=F)
AICc(owls_glmm1, owls_glmm2) %>% 
  rownames_to_column() %>%
  arrange(AICc) %>%
  mutate(dAICc = AICc - min(AICc))
```
They are nearly the same (delta AIC <2), so we can keep in the interaction term with minimal change.

```{r}
owls_glmm1a <- glmmTMB(ncalls ~ food * sex + offset(log(brood_size)) +
                         (1|nest), family = poisson(link = "log"), 
                       data = owls, REML=T)
owls_glmm1b <- glmmTMB(ncalls ~ food * sex + offset(log(brood_size)) +
                         (food|nest), family = poisson(link = "log"), 
                       data = owls, REML=T)
owls_glmm1c <- glmmTMB(ncalls ~ food * sex + offset(log(brood_size)) +
                         (sex|nest), family = poisson(link = "log"), 
                       data = owls, REML=T)
owls_glmm1d <- glmmTMB(ncalls ~ food * sex + offset(log(brood_size)) +
                         (food+sex|nest), family = poisson(link = "log"), 
                       data = owls, REML=T)
owls_glmm1e <- glmmTMB(ncalls ~ food*sex + offset(log(brood_size)) +
                         (food*sex|nest), family = poisson(link = "log"),
                       data = owls, REML=T)
AICc(owls_glmm1a, owls_glmm1b, owls_glmm1c, owls_glmm1d, owls_glmm1e) %>%
  rownames_to_column() %>%
  arrange(AICc) %>%
  mutate(dAICc = AICc - min(AICc))
```
Clearly, the last option is miles better.
This suggests that the residuals are reduced when we account for both the food and sex factor combinations, rather than one in particular!

```{r}
rm(owls_glmm1, owls_glmm2, owls_glmm1a, owls_glmm1b, owls_glmm1c, owls_glmm1d)
```

Random slopes give you more power, but will cost you degrees of freedom for no benefit where the individuals are not clearly different across treatments!

Note that it almost makes no difference which random effect structure we choose!




# Model validation {.tabset .tabset-faded}


```{r}
x <- simulateResiduals(owls_glmm1e)
testResiduals(x)
```

Clearly evidence of multiple violations, suggesting a bad model. It may be zero inflated and/or overdispersed (though the disperson is not significant at present)

```{r}
testZeroInflation(x)
performance::check_zeroinflation(owls_glmm1e)
```

Clearly it is zero-inflated!


**Conclusions:**

- there is evidence that the model does not fit that well. It is evidently zero
  inflated and possibly also overdispersed.
- it would seem that a zero-inflated Poisson or even a zero-inflated Negative  Binomial would be a sensible next step.
- zero-inflated models cannot be fit in `glmer()`, so we will proceed with `glmmTMB()` only.



# Model refit and validation {.tabset .tabset-faded}

Do we think these are false zeros or real zeros?
Did the microphones miss true calls? Possibly, yes. => zero-inflated (use ziformula without changing family)
Are the processes that generate zeros vs. other numbers different? Likely not in this particular case? => not hurdle (use ziformula AND change to truncated_#### family of models!)

```{r}
owls_glmm3 <- glmmTMB(ncalls ~ food*sex + offset(log(brood_size)) +
                         (food*sex|nest), family = poisson(link = "log"),
                       ziformula = ~1, data = owls, REML=T)

AICc(owls_glmm1e, owls_glmm3) %>% arrange(AICc)
```

Note that to specify a zero-inflated model, it is simply: ziformula = ~1
ZI much better in terms of AICc!

```{r}
x <- simulateResiduals(owls_glmm3)
testResiduals(x) # middle of QQplot looks bad, could be improved
testZeroInflation(x) # looks better
testDispersion(x) # dispersion alright
```
The middle of the QQplot looks not so good, but otherwise it appears alright.

The distribution looks almost underdispersed, meaning that when the calls were very high, its possible that the calls were not picked up upon at all! 

What this indicates is that the zeros might not be only in the low call rates, but higher call rates may have been ranked as lower, resulting in less disperson! To account for underdispersion => negative binomial

Try adding both zero-inflated functions that differ in terms of food and sex, as well as changing to a negativeb inomial
```{r}
owls_glmm4 <- glmmTMB(ncalls ~ food*sex + offset(log(brood_size)) +
                         (food*sex|nest), family = poisson(link = "log"), ziformula = ~food*sex, data = owls, REML=T)

better_opt <- glmmTMBControl(optimizer = "optim",
               optArgs = "Nelder-Mead")

# owls_glmm5 <- glmmTMB(ncalls ~ food*sex + offset(log(brood_size)) +
#                          (food*sex|nest), family = nbinom2(link = "log"), ziformula = ~1, data = owls, REML=T) # failed
# owls_glmm6 <- glmmTMB(ncalls ~ food*sex + offset(log(brood_size)) +
#                          (food*sex|nest), family = nbinom2(link = "log"), ziformula = ~food*sex, data = owls, REML=T, control = better_opt) # failed


# nbinom2 is the quadratic parameterization, but is otherwise the same, and is more useful in this format as an overdispersed poisson

AICc(owls_glmm3, owls_glmm4) %>% arrange(AICc)
```

Best model is zero-inflated with differences in detectability between the treatment combinations.

```{r}
x <- simulateResiduals(owls_glmm4, plot=T)
```



# Partial plots {.tabset .tabset-faded}

## plot_model

```{r partialPlots1a, results='markdown', eval=TRUE}
plot_model(owls_glmm4,  type='eff',  terms=c('food', 'sex')) # this is not truly correct...
```
**Note:** This plot changes our average effect sizes and multiplying by the average brood_size to get this plot, so not super useful with the offset.

These predictions appear to be based on the mean BroodSize of approximately `r round(mean(owls$BroodSize), 2)`.

## ggemmeans

`ggemmeans()` can accommodate the offset correctly.  There are two sensible
choices:

- set the offset to 0.  This results in predictions appropriate for a per owl
  chick conclusion.

```{r partialPlots1d2, results='markdown', eval=TRUE}
ggemmeans(owls_glmm4,  ~food|sex, offset=0) %>% plot
```

- set the offset to the (log) of the mean BroodSize (similar to other partial  effects), hence giving predictions appropriate for the average brood size conclusion.
  
```{r partialPlots1d1, results='markdown', eval=TRUE}
#off<-owls %>% group_by(sex, food) %>% summarize(Mean=mean(BroodSize))
off <- owls %>% summarize(Mean = mean(brood_size))
as.numeric(off)
ggemmeans(owls_glmm4,  ~food+sex, offset=log(off$Mean)) %>% plot
```

## allEffects

```{r partialPlots1b, results='markdown', eval=TRUE}
plot(allEffects(owls_glmm4), multiline=TRUE, ci.style='bars')
```

These predictions also appear to be based on the mean BroodSize, although the documentation seems to suggest that `allEffects()` might not deal with the offsets the way we have used them (as a function in the formula) correctly.

## ggpredict

```{r partialPlots1c, results='markdown', eval=TRUE}
ggpredict(owls_glmm4,  terms=c('food', 'sex')) %>% plot
```

**This seems to deal with the offset incorrectly**.  For the purpose of
prediction, the offset seems to be set at the value of the first BroodSize (on
the response scale).  This is incorrect for two reasons:

1. it should be on the log scale
2. it would be better to use either the mean BroodSize (then on the link scale)
   or a value of 0 (so as to reflect the per unit BroodSize prediction).
   

# Model investigation / hypothesis testing {.tabset .tabset-faded}

```{r}
summary(owls_glmm4)
# By count model:
#   No evidence of an interaction
#   By deprived females:
exp(0.86455) # intercept = 2.374
#   By satiated females
exp(-0.82476) # 
exp(0.86455 -0.82476) # decline in ncalls within satiated females
#   By deprived males:
exp(-0.12590)
exp(0.86455 -0.12590)

# By detection model (binomial):
#   No evidence of an interaction
#   By deprived females:
exp(-1.3762) # intercept = 0.25:1 odds of not calling correctly
plogis(-1.3762) # 20% rate of false zeros
#   By satiated females:
#   Our ability to detect calls is higher for the female satiated group
plogis(-1.3762 + 0.6988) # 34% is the rate of false zeros for the satiated
#   Our ability to detect calls is lower for the female satiated group
plogis(-1.3762 + -0.8363) # 9.9% is the rate of false zeros for the males
exp(0.6988) # 0.43 or 57% decline in ncalls within satiated group
exp(0.6988-0.8363) # 0.87 ncalls within satiated group
# by deprived males:
exp(-0.8363) # 0.43 or 57% decline in ncalls within satiated group
exp(0.6988 -0.8363) # 0.87 ncalls within satiated group
```

```{r}
# r.squaredGLMM(owls_glmm4) # will not work for zero-inflated models
performance::r2_nakagawa(owls_glmm4)
```
Note that the r^2 for GLMMs doesn't work with zero-inflated models.
Our marginal R^2 is quite low, suggesting that fixed effects weren't super important. Conditional R^2 (which includes nest and other interactions) clearly is higher and helps better explain the results!

# Predictions
```{r}
emmeans(owls_glmm4, pairwise ~ food, type = "response")
```


# Summary figures

```{r}
owls.grid <- with(owls, list(food=levels(food),
                            sex=levels(sex)))
newdata <- emmeans(owls_glmm4, ~food+sex, at=owls.grid,
                  offset=0, type='response') %>%
    as.data.frame
head(newdata)
ggplot(newdata, aes(y=rate, x=food)) +
  geom_pointrange(aes(ymin=lower.CL, ymax=upper.CL, color=sex),
                  position=position_dodge(width=0.2)) +
  scale_y_continuous('Number of sibling negotiations per chick') +
  theme_bw()

##OR if we want to express this for the average brood size
owls.grid = with(owls, list(food=levels(food),
                            sex=levels(sex)))
newdata = emmeans(owls_glmm4, ~food+sex, at=owls.grid,
                  offset=log(mean(owls$BroodSize)), type='response') %>%
    as.data.frame
head(newdata)
ggplot(newdata, aes(y=rate, x=food)) +
  geom_pointrange(aes(ymin=lower.CL, ymax=upper.CL, color=sex),
                  position=position_dodge(width=0.2)) +
  scale_y_continuous('Number of sibling negotiations per nest') +
  theme_bw()
```


```{r}

newdata=tidy(owls.glmmTMB3, effects='fixed', conf.int=TRUE,  exponentiate=TRUE) %>%
  mutate(Model='zip (simple zi)') %>%
  bind_rows(
    tidy(owls_glmm4, effects='fixed', conf.int=TRUE,  exponentiate=TRUE) %>%
    mutate(Model='zip (complex zi)')
  ) %>%
  bind_rows(
    tidy(owls.glmmTMB5, effects='fixed', conf.int=TRUE,  exponentiate=TRUE) %>%
    mutate(Model='zinb (simple zi)')
  ) %>%
  bind_rows(
    tidy(owls.glmmTMB6, effects='fixed', conf.int=TRUE,  exponentiate=TRUE) %>%
    mutate(Model='zinb (complex zi)')
  ) %>%
  mutate(Model=factor(Model,  levels=c('zip (simple zi)', 'zip (complex zi)',
                                       'zinb (simple zi)', 'zinb (complex zi)')),
         Cond=interaction(component, term)) %>%
  arrange(component, term) %>%
  mutate(Cond=factor(Cond,  levels=rev(unique(Cond))))

ggplot(newdata,  aes(y=estimate,  x=Cond,  color=Model)) +
  geom_pointrange(aes(ymin=conf.low,  ymax=conf.high),  position=position_dodge(width=0.2)) +
  coord_flip()


newdata = emmeans(owls.glmmTMB3, ~food+sex, offset=0, at=owls.grid, type='response') %>%
  as.data.frame %>% mutate(Model='zip (simple zi)',  response=rate) %>%
  bind_rows(
    emmeans(owls_glmm4, ~food+sex, offset=0, at=owls.grid, type='response') %>%
    as.data.frame %>% mutate(Model='zip (complex zi)',  response=rate)
  ) %>%
  bind_rows(
    emmeans(owls.glmmTMB5, ~food+sex, offset=0, at=owls.grid, type='response') %>%
    as.data.frame %>% mutate(Model='zinb (simple zi)',  response=response)
  ) %>%
  bind_rows(
    emmeans(owls.glmmTMB6, ~FoodTreatment+sex, offset=0, at=owls.grid, type='response') %>%
    as.data.frame %>% mutate(Model='zinb (complex zi)',  response=response)
  ) %>%
  mutate(Model=factor(Model,  levels=c('zip (simple zi)', 'zip (complex zi)',
                                       'zinb (simple zi)', 'zinb (complex zi)')))

head(newdata)
ggplot(newdata, aes(y=response, x=FoodTreatment)) +
  geom_pointrange(aes(color=sex, ymin=lower.CL, ymax=upper.CL), 
                  position=position_dodge(width=0.2)) +
  facet_wrap(~Model,  nrow=1)

```

# References

