---
title: "GLMM example 4"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(broom.mixed)
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(ggeffects) #for effects plots in ggplot
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(DHARMa)   #for residuals and diagnostics
library(nlme)     #for lme
library(lme4)      #for glmer
library(glmmTMB)    #for glmmTMB
library(performance) #for diagnostic plots
library(see)         #for diagnostic plots
theme_set(theme_classic())
```

# Scenario

To investigate synergistic coral defence by mutualist crustaceans,
@Mckeon-2012-1095 conducted an aquaria experiment in which colonies of a coral
species were placed in a tank along with a preditory seastar and one of four
symbiont combinations:

- no symbiont,
- a crab symbiont
- a shrimp symbiont
- both a crab and shrimp symbiont.

The experiments were conducted in a large octagonal flow-through seawater tank
that was partitioned into eight sections, which thereby permitted two of each of
the four symbiont combinations to be observed concurrently. The tank was left
overnight and in the morning, the presence of feeding scars on each coral colony
was scored as evidence of predation.  The experiments were repeated ten times,
each time with fresh coral colonies, seastars and symbiont.

The ten experimental times represent blocks (random effects) within which the
symbiont type (fixed effect) are nested.

# Read in the data

```{r readData, results='markdown', eval=TRUE}
mckeon <- read_csv('../data/mckeon.csv', trim_ws=TRUE) %>%
  janitor::clean_names() %>%
  mutate(
    block = factor(block),
    symbiont = fct_relevel(symbiont, c("none", "crabs", "shrimp", "both")))
glimpse(mckeon)
```


# Exploratory data analysis {.tabset .tabset-faded}

Model formula:
$$
y_i \sim{} \mathcal{N}(n, p_i)\\
ln\left(\frac{p_i}{1-p_1}\right) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of symbionts on the probability of the colony experiencing predation.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual coral colonies.


```{r eda1a, results='markdown', eval=TRUE, hidden=TRUE}
ggplot(mckeon, aes(y=predation, x=symbiont)) +
    geom_point(position=position_jitter(width=0.2, height=0))+
    facet_wrap(~block)
```

Note that we could not analyze the data if all the blocks didn't have any overlap (i.e. if all the individuals died or lived in specific circumstances)

# Fit the model {.tabset .tabset-faced}
```{r}
mckeon_glmm <- glmmTMB(predation ~ symbiont + (1|block), 
                       data = mckeon, family = binomial(link = "logit"), REML=T)
# mckeon_glmm2 <- glmmTMB(predation ~ symbiont + (symbiont|block), 
#                        data = mckeon, family = binomial(link = "logit"), REML=T) # did not run properly
```


# Model validation {.tabset .tabset-faded}
```{r}
sim_resid <- mckeon_glmm %>% simulateResiduals(plot=T)
testOverdispersion(sim_resid)
```


# Partial plots {.tabset .tabset-faded}
```{r}
plot_model(mckeon_glmm, type = 'eff')
plot(allEffects(mckeon_glmm))
ggemmeans(mckeon_glmm, ~symbiont) %>% plot(add.data=T) # marginal effects
ggpredict(mckeon_glmm) %>% plot # predictions
```


# Model investigation / hypothesis testing {.tabset .tabset-faded}

```{r}
summary(mckeon_glmm)
## For no symbiont colonies:
exp(4.420) # 83:1 odds of being predated

## For crab symbiont colonies:
exp(-3.317) # x0.03626145 fractional decrease
exp(4.420 + -3.317) # 3:1 odds of being predated
plogis(4.420 + -3.317) # probability of predation: 0.7508218

## For crab symbiont colonies:
exp(-3.317) # x0.03626145 fractional decrease
exp(4.420 + -3.317) # 3:1 odds of being predated
plogis(4.420 + -3.317) # probability of predation: 0.7508218
```
Intercept: Prob of no symbiont individuals being predated is: 83:1
Prob of crab symbiont colonies being predated declines by a factor of 0.03626145, so:
exp(4.420 + -3.317) = 3.01:1 odds of being predated
Prob of shrimp symbiont colonies being predated declines by a factor of 

```{r}

mckeon_tidy <- tidy(mckeon_glmm)
tidy(mckeon_glmm, effect='fixed', conf.int=TRUE)
tidy(mckeon_glmm, effect='fixed', conf.int=TRUE, exponentiate=TRUE)
```

**Conclusions:**

- the coefficients are presented on a logit scale.  Whilst this is not relavant
  for the purpose of inference testing, it does make it difficult to interpret
  the coefficients.
- if we exponentiate the coefficients ($log(\frac{\rho}{1-\rho})$ ->
  $\frac{\rho}{1-\rho}$), they will be presented on a odds ratio scale, and thus:
  - the intercept (none symbionts) will be 
    `r as.numeric(round(exp(mckeon_tidy[1, 5]),2))`.  That is, corals without a
    symbiont are `r as.numeric(round(exp(mckeon_tidy[1, 5]),2))` times more
    likely to be preditated on than not predated on.  The odds of predation in
    this the absence of symbionts is `r as.numeric(round(exp(mckeon_tidy[1, 5]),2))`:1.
  - in the presence of a crab symbiont, the odds of being predated on are only 
    `r as.numeric(round(exp(mckeon_tidy[2, 5]),2))` times that of the none
    symbiont group.  That is, in the presence of a crab symbiont, the odds of
    predation decline by `r 100*(1-as.numeric(round(exp(mckeon_tidy[2, 5]),2)))`%.
  - in the presence of a shrimp symbiont, the odds of being predated on are only 
    `r as.numeric(round(exp(mckeon_tidy[3, 5]),2))` times that of the none
    symbiont group.  That is, in the presence of a shrimp symbiont, the odds of
    predation decline by `r 100*(1-as.numeric(round(exp(mckeon_tidy[3, 5]),2)))`%.
  - in the presence of both crab and shrimp symbionts, the odds of being predated on are only 
    `r as.numeric(round(exp(mckeon_tidy[4, 5]),2))` times that of the none
    symbiont group.  That is, in the presence of both crab and shrimp symbiont, the odds of
    predation decline by `r 100*(1-as.numeric(round(exp(mckeon_tidy[4, 5]),2)))`%.
- if we backtransform the intercept full to the response scale (probability
  scale), ($log(\frac{\rho}{1-\rho})$ -> $\rho$), the intercept is interpreted
  as the probability that corals will be predated in the absence of of symbionts
  is `r round(plogis(as.numeric(mckeon_tidy[1, 5]),2))`
  
```{r}
emmeans(mckeon_glmm, ~ symbiont) %>% 
  as.data.frame() %>%
  mutate(emmean = exp(emmean))

mckeon_glmm %>% 
  emmeans(~ symbiont) %>% # get differences
  regrid() #%>%  # does backtransform of the differences BEFORE pair-wise comparisons
  # pairs() %>% # get pair-wise absolute values of differences
  # confint()
```

# Planned contrast

Can only do 3 planned contrasts, as there are 4 groups, so k-1 possible to evaluate.

Order is: "none", "crabs", "shrimp", "both" 

## Crab vs. shrimp
0, 1, -1, 0

## One symbiont vs. two symbiont
0, 0.5, 0.5, -1

## No symbiont vs. symbiont types
1, -1/3, -1/3, -1/3

```{r}

cmat <- cbind('crab_vs_shrimp' = c(0, 1, -1, 0),
              'one_vs_twosym' = c(0, 0.5, 0.5, -1),
              'nosym_vs_sym' = c(1, -1/3, -1/3, -1/3))
cmat
colSums(cmat) # need to sum to zero
crossprod(cmat) # needs to have only zeros on the off-diagonal (orthogonal)
```


```{r}
mckeon_glmm %>%
  emmeans(~symbiont, contr = list(symbiont = cmat), type = 'response') #'ratio'  is on the log scale
mckeon_glmm %>%
  emmeans(~symbiont, type = 'response') %>%
  contrast(list(cmat)) # returns just the table
```

Clearly there is only a difference between the no-symbiont vs. symbiont groups. 

Note that the odds ratio here is an odds ratio of the groups' odds ratios (so the odds of a colony with no symbionts being predated is 63 times higher than the odds for colonies with any type of symbiont)

```{r}
# mckeon_glmm %>%
#   emmeans(~symbiont) %>%
#   regrid() %>%
#   contrast(list(symbiont = cmat)) %>%
#   confint()
```

R^2:
```{r}
r.squaredGLMM(mckeon_glmm)
performance::r2_nakagawa(mckeon_glmm)
```

Delta method of R^2 can be used for any models. But there are some select distributions that R^2 theoretical applies to, so usually we just use whichever ones the nakagawa function provides.
Just don't use theoretical for lognormal(gaussian with a log-link) or gamma(log-link) distributions.

Only about 15% for marginal (fixed effects only) and 86% for conditional R^2, so clearly a lot is explained by individual colony variation in predation!

### tab_model

```{r summary2c, results='markdown', eval=TRUE, hidden=TRUE}
# warning this is only appropriate for html output
# sjPlot::tab_model(mckeon_glmm, show.se=TRUE, show.aic=TRUE)
```


# Summary figure {.tabset .tabset-faded}

```{r summaryFig2a, results='markdown', eval=TRUE, hidden=TRUE}
emmeans(mckeon_glmm, ~symbiont, type='response') %>% 
  as.data.frame %>%
  ggplot(aes(y=prob,  x=symbiont)) +
  geom_pointrange(aes(ymin=lower.CL,  ymax=upper.CL))
```
# References
